---
title: "Homework 5"
subtitle: "Research Methods, Spring 2025"
author: "Genevieve DeBell"
format:
  pdf:
    output-file: "debell-g-hwk5-2"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| include: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, 
               data.table, gdata, scales, kableExtra, modelsummary, fixest)

# Force modelsummary to use LaTeX it can actually render
options(modelsummary_factory_default = "kableExtra")
```


```{r}
#| include: false
#| eval: true
setwd("/Users/genevievedebell/Documents/GitHub/hwk5")
load("submission3/analysis/Hwk5_workspace.Rdata")
```
My analysis code can be found at my GitHub repository, linked [here.](https://github.com/genevievedebell/hwk5)

\noindent 1. Plot the share of the adult population with direct purchase health insurance over time.
```{r, warning=FALSE}
#| echo: false
library(ggplot2)
library(dplyr)

# Calculate the share of direct-purchase insurance
plot_data <- final.insurance %>%
  group_by(year) %>%
  summarise(
    total_direct = sum(ins_direct, na.rm = TRUE),
    total_adult_pop = sum(adult_pop, na.rm = TRUE)
  ) %>%
  mutate(direct_share = total_direct / total_adult_pop)

# Plot it
ggplot(plot_data, aes(x = year, y = direct_share)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Share of Adult Population with Direct Purchase Insurance (2012–2019)",
       x = "Year",
       y = "Direct Purchase Insurance Share") +
  theme_minimal()
```
\newpage
\noindent 2. Discuss the reduction in direct purchase health insurance in later years. Can you list a couple of policies that might have affected the success of the direct purchase insurance market?

Since 2016, the share of adults with direct purchase insurance has decreased. The Tax Cuts and Jobs Act of 2017 eliminated the penalty for not having insurance starting in 2019. Without this mandate, fewer healthy individuals opted into coverage because there was no penalty. 

\noindent 3. Plot the share of the adult population with Medicaid over time.
```{r,warning=FALSE}
#| echo: false
# Calculate Medicaid share nationally by year
medicaid_plot_data <- final.insurance %>%
  group_by(year) %>%
  summarise(
    total_medicaid = sum(ins_medicaid, na.rm = TRUE),
    total_adult_pop = sum(adult_pop, na.rm = TRUE)
  ) %>%
  mutate(medicaid_share = total_medicaid / total_adult_pop)

# Plot
ggplot(medicaid_plot_data, aes(x = year, y = medicaid_share)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Share of Adult Population with Medicaid Coverage (2012–2019)",
       x = "Year",
       y = "Medicaid Coverage Share") +
  theme_minimal()
```
\newpage
\noindent 4. Plot the share of uninsured over time, separately by states that expanded Medicaid in 2014 versus those that did not. Drop all states that expanded after 2014.
```{r, warning=FALSE}
#| echo: false
### states that expanded in 2014
expanded <- final.data %>%
  group_by(State) %>%
  summarize(first_expand_year = unique(year(date_adopted))) %>%
  mutate(
    expand_group = case_when(
      is.na(first_expand_year) ~ "Never Expanded",
      first_expand_year == 2014 ~ "Expanded in 2014",
      TRUE ~ NA_character_  # Drop states with other expansion years
    )
  ) %>%
  filter(!is.na(expand_group))

### join to main data
final.data.exp <- final.data %>%
  inner_join(expanded, by = "State")

### calculate uninsured share by year and expansion group
uninsured.share <- final.data.exp %>%
  group_by(year, expand_group) %>%
  summarize(
    total_uninsured = sum(uninsured, na.rm = TRUE),
    total_adult_pop = sum(adult_pop, na.rm = TRUE),
    share_uninsured = total_uninsured / total_adult_pop,
    .groups = "drop"
  )

### plot
uninsured.plot <- ggplot(uninsured.share, aes(x = year, y = share_uninsured, color = expand_group)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Uninsured Rate by Medicaid Expansion Status (2012–2019)",
    x = "Year",
    y = "Share Uninsured",
    color = "Expansion Status"
  ) +
  theme_minimal()

print(uninsured.plot)
```
\newpage
\noindent 5. Calculate the average percent of uninsured individuals in 2012 and 2015, separately for expansion and non-expansion states. Present your results in a basic 2x2 DD table.
```{r, warning=FALSE}
#| echo: false
library(dplyr)
library(tidyr)
library(knitr)

# Start from final.data.exp and compute only what's needed
dd_2x2_clean <- final.data.exp %>%
  filter(year %in% c(2012, 2015)) %>%
  group_by(expand_group, year) %>%
  summarise(
    avg_uninsured_rate = sum(uninsured, na.rm = TRUE) / sum(adult_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = avg_uninsured_rate,
    names_prefix = "Year_"
  ) %>%
  rename(Group = expand_group, Pre = Year_2012, Post = Year_2015)

# Print the final 2x2 table
kable(dd_2x2_clean,
      caption = "Table 1: Difference-in-Differences Table of Average Uninsured Rate",
      digits = 2,
      format = "latex",
      booktabs = TRUE)

```
\newpage
\noindent 6. Estimate the effect of Medicaid expansion on the uninsurance rate using a standard DD regression estimator, again focusing only on states that expanded in 2014 versus those that never expanded.
```{r}
#| echo: false
#| warning: false
#| label: tbl-ddregs
#| tbl-cap: "DD Estimates for Medicaid Expansion"

# Create treatment and post indicators
final.data.exp <- final.data.exp %>%
  mutate(
    treat = ifelse(expand_group == "Expanded in 2014", 1, 0),
    post = ifelse(year >= 2014, 1, 0),
    uninsured_rate = uninsured / adult_pop
  )

# Estimate DiD model
did_model <- feols(uninsured_rate ~ treat * post, data = final.data.exp)

# Render formatted table
modelsummary(
  did_model,
  output = "kableExtra",
  title = "Table 2: DD Estimates for Medicaid Expansion",
  coef_rename = c(
    "treat" = "Treatment Group",
    "post" = "Post-2014",
    "treat:post" = "Interaction (Treatment × Post)"
  ),
  statistic = "({std.error})",
  stars = TRUE,
  gof_omit = "DF|AIC|BIC|Log|Adj"
)
```
\newpage
\noindent 7. Include state and year fixed effects in your estimates. Try using the lfe or fixest package to estimate this instead of directly including the fixed effects.
```{r}
#| echo: false
#| warning: false
#| label: tbl-ddregs2
#| tbl-cap: "DD Estimates with State and Year Fixed Effects"

# Fixed effects model
did_fe_model <- feols(
  uninsured_rate ~ treat * post | State + year,
  data = final.data.exp
)

# Format side-by-side table
models_fe <- list("Standard DD" = did_model, "TWFE" = did_fe_model)

modelsummary(
  models_fe,
  output = "kableExtra",
  title = "Table 3: DD with State & Year Fixed Effects",
  coef_rename = c(
    "treat" = "Treatment Group",
    "post" = "Post-2014",
    "treat:post" = "Interaction (Treatment × Post)"
  ),
  statistic = "({std.error})",
  stars = TRUE,
  gof_omit = "DF|AIC|BIC|Log|Adj"
)
```
\newpage
\noindent 8. Repeat the analysis in question 7 but include all states (even those that expanded after 2014). Are your results different? If so, why?

Yes, the results differ slightly when we include all states, including those that expanded Medicaid after 2014. This is because the timing of Medicaid expansion varies across states. States that expanded later may have had different baseline uninsurance rates, economic conditions, or implementation processes, all of which can influence the estimated average treatment effect. By accounting for staggered treatment timing, the model captures a time-varying impact of Medicaid expansion compared to a simple pre/post framework.


```{r}
#| echo: false
#| warning: false
#| label: tbl-ddregs3
#| tbl-cap: "DD Estimates with Staggered Treatment Timing"

# Create time-varying treatment variable
reg.data2 <- final.data %>%
  mutate(
    treat = case_when(
      year >= expand_year & !is.na(expand_year) ~ 1,
      is.na(expand_year) ~ 0,
      year < expand_year ~ 0
    ),
    perc_unins = uninsured / adult_pop
  )

# Estimate DiD model with FE
fe.est2 <- feols(perc_unins ~ treat | State + year, data = reg.data2)

# Combine into one table
models_all <- list(
  "Standard DD" = did_model,
  "TWFE" = did_fe_model,
  "Staggered Treatment" = fe.est2
)

modelsummary(
  models_all,
  output = "kableExtra",
  title = "Table 4: DD with Staggered Expansion",
  coef_rename = c("treat" = "Treatment Effect"),
  statistic = "({std.error})",
  stars = TRUE,
  gof_omit = "DF|AIC|BIC|Log|Adj"
)
```

\newpage
\noindent 9. Provide an “event study” graph showing the effects of Medicaid expansion in each year. Use the specification that includes state and year fixed effects, limited to states that expanded in 2014 or never expanded.
```{r, warning=FALSE}
#| echo: false
reg.data <- final.data %>%
  filter(expand_year == 2014 | is.na(expand_year)) %>%
  mutate(
    expand_ever = ifelse(!is.na(expand_year), 1, 0),
    perc_unins = uninsured / adult_pop
  )

mod.twfe <- feols(
  perc_unins ~ i(year, expand_ever, ref = 2013) | State + year,
  cluster = ~State,
  data = reg.data
)

# Plot coefficients
iplot(mod.twfe,
      main = "Event Study: Effect of Medicaid Expansion on Uninsurance",
      xlab = "Year (Reference = 2013)",
      ylab = "Effect on % Uninsured")
```
\newpage
\noindent 10. Repeat part 9 but again include states that expanded after 2014. Note: this is tricky…you need to put all states onto “event time” to create this graph.
```{r, warning=FALSE}
#| echo: false
reg.data2 <- final.data %>%
  filter(!is.na(year)) %>%
  mutate(
    expand_ever = ifelse(!is.na(expand_year), TRUE, FALSE),
    perc_unins = uninsured / adult_pop,
    time_to_treat = ifelse(expand_ever == TRUE, year - expand_year, -1),
    time_to_treat = ifelse(time_to_treat <= -4, -4, time_to_treat)
  )

library(fixest)

mod.twfe2 <- feols(
  perc_unins ~ i(time_to_treat, expand_ever, ref = -1) | State + year,
  cluster = ~State,
  data = reg.data2
)

iplot(mod.twfe2,
      main = "Event Study: Medicaid Expansion (All States, Event Time)",
      xlab = "Years Since Expansion",
      ylab = "Effect on % Uninsured")
```