---
title: "Homework 5"
subtitle: "Research Methods, Spring 2025"
author: "Answer Key"
format:
  pdf:
    output-file: "debell-g-hwk5-2"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| include: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, 
               data.table, gdata, scales, kableExtra, modelsummary, fixest)
```


```{r}
#| include: false
#| eval: true
setwd("/Users/genevievedebell/Documents/GitHub/hwk5")
load("submission2/analysis/Hwk5_workspace.Rdata")
```

1. Plot the share of the adult population with direct purchase health insurance over time.
```{r, warning=FALSE}
#| echo: false
library(ggplot2)
library(dplyr)

# Calculate the share of direct-purchase insurance
plot_data <- final.insurance %>%
  group_by(year) %>%
  summarise(
    total_direct = sum(ins_direct, na.rm = TRUE),
    total_adult_pop = sum(adult_pop, na.rm = TRUE)
  ) %>%
  mutate(direct_share = total_direct / total_adult_pop)

# Plot it
ggplot(plot_data, aes(x = year, y = direct_share)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Share of Adult Population with Direct Purchase Insurance (2012–2019)",
       x = "Year",
       y = "Direct Purchase Insurance Share") +
  theme_minimal()
```

2. Discuss the reduction in direct purchase health insurance in later years. Can you list a couple of policies that might have affected the success of the direct purchase insurance market?

Since 2016, the share of adults with direct purchase insurance has decreased. The Tax Cuts and Jobs Act of 2017 eliminated the penalty for not having insurance starting in 2019. Without this mandate, fewer healthy individuals opted into coverage because there was no penalty. 

3. Plot the share of the adult population with Medicaid over time.
```{r,warning=FALSE}
#| echo: false
# Calculate Medicaid share nationally by year
medicaid_plot_data <- final.insurance %>%
  group_by(year) %>%
  summarise(
    total_medicaid = sum(ins_medicaid, na.rm = TRUE),
    total_adult_pop = sum(adult_pop, na.rm = TRUE)
  ) %>%
  mutate(medicaid_share = total_medicaid / total_adult_pop)

# Plot
ggplot(medicaid_plot_data, aes(x = year, y = medicaid_share)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Share of Adult Population with Medicaid Coverage (2012–2019)",
       x = "Year",
       y = "Medicaid Coverage Share") +
  theme_minimal()
```

4. Plot the share of uninsured over time, separately by states that expanded Medicaid in 2014 versus those that did not. Drop all states that expanded after 2014.
```{r, warning=FALSE}
#| echo: false
### states that expanded in 2014
expanded <- final.data %>%
  group_by(State) %>%
  summarize(first_expand_year = unique(year(date_adopted))) %>%
  mutate(
    expand_group = case_when(
      is.na(first_expand_year) ~ "Never Expanded",
      first_expand_year == 2014 ~ "Expanded in 2014",
      TRUE ~ NA_character_  # Drop states with other expansion years
    )
  ) %>%
  filter(!is.na(expand_group))

### join to main data
final.data.exp <- final.data %>%
  inner_join(expanded, by = "State")

### calculate uninsured share by year and expansion group
uninsured.share <- final.data.exp %>%
  group_by(year, expand_group) %>%
  summarize(
    total_uninsured = sum(uninsured, na.rm = TRUE),
    total_adult_pop = sum(adult_pop, na.rm = TRUE),
    share_uninsured = total_uninsured / total_adult_pop,
    .groups = "drop"
  )

### plot
uninsured.plot <- ggplot(uninsured.share, aes(x = year, y = share_uninsured, color = expand_group)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Uninsured Rate by Medicaid Expansion Status (2012–2019)",
    x = "Year",
    y = "Share Uninsured",
    color = "Expansion Status"
  ) +
  theme_minimal()

print(uninsured.plot)
```

5. Calculate the average percent of uninsured individuals in 2012 and 2015, separately for expansion and non-expansion states. Present your results in a basic 2x2 DD table.
```{r, warning=FALSE}
#| echo: false
library(dplyr)
library(tidyr)


# Start from final.data.exp and compute only what's needed
dd_2x2_clean <- final.data.exp %>%
  filter(year %in% c(2012, 2015)) %>%
  group_by(expand_group, year) %>%
  summarise(
    avg_uninsured_rate = sum(uninsured, na.rm = TRUE) / sum(adult_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = avg_uninsured_rate,
    names_prefix = "Year_"
  ) %>%
  rename(Group = expand_group, Pre = Year_2012, Post = Year_2015)

# Print the final 2x2 table
kable(dd_2x2_clean,
      caption = "Table 1: Difference-in-Differences Table of Average Uninsured Rate",
      digits = 2,
      format = "latex",
      booktabs = TRUE)

```

6. Estimate the effect of Medicaid expansion on the uninsurance rate using a standard DD regression estimator, again focusing only on states that expanded in 2014 versus those that never expanded.
```{r, warning=FALSE}
#| echo: false
library(fixest)
library(dplyr)

# Create treatment and post indicators
final.data.exp <- final.data.exp %>%
  mutate(
    treat = ifelse(expand_group == "Expanded in 2014", 1, 0),
    post = ifelse(year >= 2014, 1, 0),
    uninsured_rate = uninsured / adult_pop
  )

  # DiD regression without fixed effects
did_model <- feols(uninsured_rate ~ treat * post, data = final.data.exp)
library(modelsummary)
modelsummary(
  did_model,
  output = "latex",
  title = "Table 2: DD Estimates for Medicaid Expansion",
  stars = TRUE,
  statistic = "({std.error})"
)
```

7. Include state and year fixed effects in your estimates. Try using the lfe or fixest package to estimate this instead of directly including the fixed effects.
```{r, warning=FALSE}
#| echo: false
library(fixest)

# Estimate DiD model with state and year fixed effects
did_fe_model <- feols(
  uninsured_rate ~ treat * post | State + year,
  data = final.data.exp
)

# View the summary
summary(did_fe_model)
```

8. Repeat the analysis in question 7 but include all states (even those that expanded after 2014). Are your results different? If so, why?

```{r, warning=FALSE}
#| echo: false
library(fixest)
library(dplyr)

# Step 1: Define treatment as starting in each state's actual expansion year
reg.data2 <- final.data %>%
  mutate(
    treat = case_when(
      year >= expand_year & !is.na(expand_year) ~ 1,     # State has expanded by this year
      is.na(expand_year) ~ 0,                            # State never expands
      year < expand_year & !is.na(expand_year) ~ 0       # State will expand, but not yet
    ),
    perc_unins = uninsured / adult_pop
  )

# Step 2: Estimate DiD model with state and year fixed effects
fe.est2 <- feols(
  perc_unins ~ treat | State + year,
  data = reg.data2
)

# Step 3: View results
summary(fe.est2)
```

9. Provide an “event study” graph showing the effects of Medicaid expansion in each year. Use the specification that includes state and year fixed effects, limited to states that expanded in 2014 or never expanded.
```{r, warning=FALSE}
#| echo: false
reg.data <- final.data %>%
  filter(expand_year == 2014 | is.na(expand_year)) %>%
  mutate(
    expand_ever = ifelse(!is.na(expand_year), 1, 0),
    perc_unins = uninsured / adult_pop
  )

mod.twfe <- feols(
  perc_unins ~ i(year, expand_ever, ref = 2013) | State + year,
  cluster = ~State,
  data = reg.data
)

# Plot coefficients
iplot(mod.twfe,
      main = "Event Study: Effect of Medicaid Expansion on Uninsurance",
      xlab = "Year (Reference = 2013)",
      ylab = "Effect on % Uninsured")
```

# 10. Repeat part 9 but again include states that expanded after 2014. Note: this is tricky…you need to put all states onto “event time” to create this graph.
```{r, warning=FALSE}
#| echo: false
reg.data2 <- final.data %>%
  filter(!is.na(year)) %>%
  mutate(
    expand_ever = ifelse(!is.na(expand_year), TRUE, FALSE),
    perc_unins = uninsured / adult_pop,
    time_to_treat = ifelse(expand_ever == TRUE, year - expand_year, -1),
    time_to_treat = ifelse(time_to_treat <= -4, -4, time_to_treat)
  )

library(fixest)

mod.twfe2 <- feols(
  perc_unins ~ i(time_to_treat, expand_ever, ref = -1) | State + year,
  cluster = ~State,
  data = reg.data2
)

iplot(mod.twfe2,
      main = "Event Study: Medicaid Expansion (All States, Event Time)",
      xlab = "Years Since Expansion",
      ylab = "Effect on % Uninsured")
```